Proving linearizability using forward simulations\thanks{An extended version of this paper including the missing proofs can be found at \cite{extended}.}

### 2017-05-05 draftCAV/queue.tex
Separating stacked figures.
Figure separation.
Locations in the abstract state machine rules.

\begin{figure}[t]
  \begin{minipage}[b]{0.55\linewidth}
    \begin{lstlisting}
void enq(int x) {
  i = back++; items[i] = x;
}
int deq() {
  while (1) {
    range = back - 1;
    for (int i = 0; i <= range; i++) {
      x = swap(items[i],null);
      if (x != null) return x;
} } }
    \end{lstlisting}
    \caption{The Herlihy \& Wing Queue~\cite{journals/toplas/HerlihyW90}.}
    \label{fig:HerlihyWing}
    \vspace{1em}

    \includegraphics[width=7cm]{fig-queue1}
    \includegraphics[width=7cm]{fig-queue2}
    \caption{Forward simulation with $AbsQ$. Lines depict operations,
    and circles depict call, return, and linearization point actions.}
    \label{fig:queueSim}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.4\linewidth}
    \begin{lstlisting}
function present: $\mathbb{O} \to \mathbb{B}$
function pending: $\mathbb{O} \to \mathbb{B}$
function arg: $\mathbb{V} \to \mathbb{O}$
function before: $\mathbb{O} \times \mathbb{O} \to \mathbb{B}$

rule inv(enq,v):
  let k := new $\mathbb{O}$
  present(k) := true
  pending(k) := true
  arg(v) := k
  forall k' with present(k'):
    if $\lnot$pending(k'):
      before(k',k) := true

rule lin(enq,v):
  pass

rule ret(enq,v):
  let k := arg(v)
  pending(k) := false

rule inv(deq,v):
  pass

rule lin(deq,v):
  if v = EMPTY:
    forall k with present(k):
      assert pending(k)
  else:
    let k := arg(v)
    assert present(k)
    forall k' with present(k'):
      assert $\lnot$before(k', k)
    present(k) := false

rule ret(deq,v):
  pass
    \end{lstlisting}
    \caption{The $AbsQ$ implementation.}
    \label{fig:transitions:AbsQ}
  \end{minipage}
\end{figure}

>>>>

\begin{figure}[t]
  \begin{minipage}[b]{0.55\linewidth}
    \begin{lstlisting}
void enq(int x) {
  i = back++; items[i] = x;
}
int deq() {
  while (1) {
    range = back - 1;
    for (int i = 0; i <= range; i++) {
      x = swap(items[i],null);
      if (x != null) return x;
} } }
    \end{lstlisting}
    \caption{The Herlihy \& Wing Queue~\cite{journals/toplas/HerlihyW90}.}
    \label{fig:HerlihyWing}

    \smallskip
    \hrule
    \smallskip

    \includegraphics[width=7cm]{fig-queue1}
    \includegraphics[width=7cm]{fig-queue2}
    \caption{Forward simulation with $AbsQ$. Lines depict operations,
    and circles depict call, return, and linearization point actions.}
    \label{fig:queueSim}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.38\linewidth}
    \begin{lstlisting}
function loc: $\mathbb{O} \to \set{\tt inv, lin, ret, \bot}$
function arg, ret: $\mathbb{O} \to \mathbb{V}$
function present, pending: $\mathbb{O} \to \mathbb{B}$
function before: $\mathbb{O} \times \mathbb{O} \to \mathbb{B}$

rule inv(enq,v,k):
  arg(k) := v
  present(k) := true
  pending(k) := true
  forall k1 with present(k1):
    if $\lnot$pending(k1):
      before(k1,k) := true

rule ret(enq,k):
  pending(k) := false

rule inv(deq,k):
  pass

rule lin(deq,v,k):
  ret(k) := v
  if v = EMPTY:
    forall k' with present(k'):
      assert pending(k')
  else:
    let k1 = arg$^{-1}$(v)
    assert present(k1)
    forall k2 with present(k2):
      assert $\lnot$before(k2,k1)
    present(k1) := false

rule ret(deq,v,k):
  assert ret(k) = v
    \end{lstlisting}
    \caption{The $AbsQ$ implementation; each rule \lstinline|$\alpha$(\_,k)|
    implicitly begins with \lstinline|assert loc(k)=$\alpha$| and ends with
    the appropriate \lstinline|loc(k):=$\beta$|.}
    \label{fig:transitions:AbsQ}
  \end{minipage}
\end{figure}
